---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", export_all = FALSE, helpers = FALSE, quiet = TRUE)
} else {
  library(tccquickr)
}
knitr::knit_engines$set(rtinycc = Rtinycc:::rtinycc_engine)

tccq_c_block <- function(code) {
  Rtinycc:::rtinycc_c_block(code)
}
```

# tccquickr

Experimental backend-neutral R code transformation framework on top of
`Rtinycc`.

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/tccquickr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/tccquickr/actions/workflows/R-CMD-check.yaml)
[![tccquickr status badge](https://sounkou-bioinfo.r-universe.dev/tccquickr/badges/version)](https://sounkou-bioinfo.r-universe.dev/tccquickr)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

## Abstract

`tccquickr` is an experimental compiler and transformation framework for a
small, declared R subset.

The current design direction is centered on the `tccq_*` path:

- frontend parsing and typed lowering for `declare(type(...))`-annotated R
- explicit middle-end IR for producers, materialization, folds, statements, and
  legality barriers
- C emission through the R C API
- swappable C backends, with source-only output, TinyCC via
  [`Rtinycc`](https://github.com/sounkou-bioinfo/Rtinycc), and shared-library
  compilation through `R CMD SHLIB`, in the same general deployment space as
  [`callme`](https://github.com/coolbutuseless/callme)

The package depends on [`Rtinycc`](https://github.com/sounkou-bioinfo/Rtinycc)
for the underlying TinyCC toolchain, libtcc runtime, FFI compilation pipeline,
and pointer/runtime support.

Conceptually, the project sits near several adjacent efforts:

- [`quickr`](https://github.com/t-kalinowski/quickr) as a useful R subset
  compiler comparison
- [`anvil`](https://github.com/r-xla/anvil) as a broader transformation and
  backend framework comparison
- [`callme`](https://github.com/coolbutuseless/callme) as a useful reference
  for C-only shared-library compilation/loading workflows
- [`SAC` / `sac2c`](https://sac-home.org/about%3Asac) as the main inspiration
  for explicit array IR, fusion, and allocation reduction

## Scope

`tccquickr` is where the transformation framework lives.

The main moving parts are:

- `tccq_compile()`
- backend factories such as `tccq_backend_source()`,
  `tccq_backend_tinycc()`, and `tccq_backend_shlib()`
- typed frontend parsing and lowering
- middle-end kernel IR and rewrite passes
- C + R C API target emission
- tinytest coverage for generated code and compiled behavior

[`Rtinycc`](https://github.com/sounkou-bioinfo/Rtinycc) remains responsible for:

- bundled TinyCC build and installation
- `tcc_ffi()`, `tcc_bind()`, and `tcc_compile()`
- low-level state and CLI helpers
- pointer, struct, callback, and declarative FFI support

Shared-library deployment through `R CMD SHLIB` is a separate C backend path in
`tccquickr`, closer in spirit to [`callme`](https://github.com/coolbutuseless/callme)
than to `Rtinycc`'s TinyCC runtime responsibilities.

## Installation

```{r eval = FALSE}
install.packages(
  "tccquickr",
  repos = c(
    "https://sounkou-bioinfo.r-universe.dev",
    "https://cloud.r-project.org"
  )
)
```

## Current `tccq` compiler

`tccq_*` is the current compiler architecture.

The main design rule is that the compiler knows more while it is still working
with typed R-level IR than after it has emitted C. So legality, alias/view
semantics, mutation barriers, fallback boundaries, and materialization choices
should live in the frontend or middle-end whenever possible, with the C target
mostly consuming those decisions.

### Current status

Implemented now:

- [x] declared scalar and vector inputs
- [x] `double`, `integer`, and `logical` typing
- [x] scalar arithmetic
- [x] vector elementwise arithmetic
- [x] comparison and logical vector expressions
- [x] unary math calls such as `sin()`, `cos()`, `exp()`, `log()`, and `sqrt()`
- [x] generic fold-style reducers: `sum`, `prod`, `min`, `max`, `mean`, `any`, and `all`
- [x] explicit kernel IR nodes for `domain`, `producer`, `materialize`, `fold`, and `scalar_kernel`
- [x] statement/program IR for local bindings and writes
- [x] explicit contiguous view IR via `view1`
- [x] scalar indexed reads `x[i]`
- [x] contiguous range slices `x[lo:hi]`
- [x] local indexed writes `y[i] <- v` and local range writes `y[lo:hi] <- v`
- [x] limited `Reduce(FUN, x)` lowering for recognized reducer surfaces
- [x] explicit fallback boundary nodes in `fallback = "auto"`
- [x] conservative storage, allocation, and protect planning passes
- [x] compile-time backend/target capability checks for context fields and boundary APIs
- [x] source-only, TinyCC-backed, and shared-library (`R CMD SHLIB`) backends
- [x] generated differential validation over programmatically constructed cases

Planned next:

- [ ] richer semantic allocation/reuse planning before C emission
- [ ] clearer middle-end ownership of boundary argument materialization
- [ ] more systematic view/index normalization before target codegen
- [ ] axis-wise reductions and matrix-aware lowering
- [ ] broader indexing forms such as gather/scatter/filter
- [ ] larger harvested validation corpus in addition to generated tests

### Current architecture split

- [x] frontend: parse `declare(type(...))` and lower a small R subset
- [x] middle-end: validate IR, infer effects, kernelize, lower recognized reducers, fuse, handle boundaries, and derive conservative plans
- [x] target: emit C using the R C API
- [x] backend: return source or compile emitted C through TinyCC or `R CMD SHLIB`
- [ ] middle-end: richer plan-driven allocation/reuse and view/index normalization
- [ ] backend expansion toward other C compilation/loading paths such as additional system compiler or `callme`-style workflows

### Reducer and idiom checklist

Implemented now:

- [x] direct reducers `sum`, `prod`, `min`, `max`, `mean`, `any`, `all`
- [x] vector comparisons feeding reducers, e.g. `any(x > 0)`
- [x] vector logical composition feeding reducers, e.g. `all((x > 0) & (y > 0))`
- [x] reducer lowering through explicit `fold` kernels rather than `sum`-only special casing
- [x] limited `Reduce(FUN, x)` lowering for recognized reducer operators such as `+`, `*`, `&`, and `|`

Still open:

- [ ] full base-R `Reduce()` surface compatibility
- [ ] axis-wise reducers such as `rowSums()` / `colSums()`
- [ ] broader `apply`-family lowering on top of rank-aware IR

### Validation checklist

Implemented now:

- [x] explicit hand-written runtime tests for reducers, slices, views, boundaries, and backends
- [x] generated differential tests comparing compiled `tccq` code against direct R evaluation
- [x] cross-backend checks between TinyCC and `R CMD SHLIB`

Still open:

- [ ] larger maintained validation corpus harvested from real array-oriented R code
- [ ] pass-by-pass translation-validation style checks for more middle-end rewrites
- [ ] broader edge-case matrices around exact reducer warning/NA behavior

### Backend selection

`tccq_compile()` accepts explicit backend objects:

- `tccq_backend_source()` returns emitted C without compiling it
- `tccq_backend_tinycc()` compiles emitted C in memory through `Rtinycc`
- `tccq_backend_shlib()` compiles emitted C through `R CMD SHLIB`

That shared-library path is in the same general deployment space as
[`callme`](https://github.com/coolbutuseless/callme): a C-only compile/load
workflow around `.Call()` entry points. Internally, `tccq_compile()` validates
backend capabilities against the target, compile context, and explicit boundary
APIs before compilation.

### Kernel IR concepts

`producer`
: an element computation over an index domain

`materialize(producer)`
: allocate an output vector and write producer elements into it

`fold(op, domain, elem)`
: reduce a producer-like element expression over a loop domain

These are the concepts that let the package move from “fusion hidden in code
emission” toward “fusion represented in IR”.

### Assignment and slicing rules

The current compiler treats these forms differently:

- `a <- expr` is a local binding, not mutation
- `x[i]` is an indexed read with checked 1-based indexing
- `x[lo:hi]` is a contiguous slice expression
- `a[i] <- v` and `a[lo:hi] <- v` are mutation barriers

For the current milestone, indexed writes are only allowed into local vector
bindings. Direct formal mutation such as `x[i] <- v` is rejected.

### Example: inspect fresh IR

```{r}
fresh_sum_kernel <- function(x, y) {
  declare(type(x = double(NA), y = double(NA)))
  sum((sin(x) + y) * y)
}

fresh_vec_kernel <- function(x, y) {
  declare(type(x = double(NA), y = double(NA)))
  sin(x) + y * y
}

sum_module <- tccq_compile(fresh_sum_kernel, mode = "ir")
vec_module <- tccq_compile(fresh_vec_kernel, mode = "ir")

sum_module
vec_module
```

The reduction path is represented as a `fold` over an explicit `producer`.
The vector return path is represented as `materialize(producer)`.

### Example: explicit fusion rewrite

Before fusion, `kernelize` represents the reduction as a fold over a
materialized producer. The fusion pass rewrites that to a fold over the
producer directly.

```{r}
mod0 <- tccquickr:::tccq_frontend(fresh_sum_kernel)
mod0 <- tccquickr:::tccq_lower_module(mod0)

before_fusion <- tccquickr:::tccq_run_passes(
  mod0,
  passes = list(
    tccquickr:::tccq_pass_validate_ir(),
    tccquickr:::tccq_pass_effects(),
    tccquickr:::tccq_pass_kernelize()
  )
)

after_fusion <- tccquickr:::tccq_run_passes(
  mod0,
  passes = list(
    tccquickr:::tccq_pass_validate_ir(),
    tccquickr:::tccq_pass_effects(),
    tccquickr:::tccq_pass_kernelize(),
    tccquickr:::tccq_pass_fusion()
  )
)

before_fusion
after_fusion
```

This is still a small rewrite system, but the important point is that fusion is
now expressed as a middle-end transformation rather than being buried entirely
inside C string emission.

### Example: explicit boundary fallback

Boundary nodes stay explicit legality barriers, but `fallback = "auto"` now
lowers unsupported calls to explicit `r_eval` boundary nodes instead of failing
immediately during lowering.

```{r}
fallback_kernel <- function(x) {
  declare(type(x = double()))
  floor(x)
}

fallback_ir <- tccq_compile(fallback_kernel, mode = "ir", fallback = "auto")
fallback_ir$ir$result$tag
fallback_ir$ir$result$api
```

In `fallback = "hard"` mode, the same unsupported call is rejected.

```{r}
tryCatch(
  tccq_compile(fallback_kernel, mode = "ir", fallback = "hard"),
  error = function(e) e$message
)
```

### Example: assignments and indexed writes

```{r}
assign_kernel <- function(x, i, v) {
  declare(type(x = double(NA)), type(i = integer()), type(v = double()))
  y <- x
  y[i] <- v
  y
}

assign_src <- tccq_compile(assign_kernel, mode = "code")
tccq_c_block(assign_src)
```

The local `y <- x` binding now starts as an alias to the formal input. The
first indexed write materializes `y` into owned local storage, so mutate-then-
return allocates only once while still keeping direct writes to caller-owned
formal vectors out of scope for this milestone.

### Example: slicing and reduction

```{r}
slice_sum_kernel <- function(x, lo, hi) {
  declare(type(x = double(NA)), type(lo = integer()), type(hi = integer()))
  y <- x[lo:hi]
  sum(y)
}

slice_sum_src <- tccq_compile(slice_sum_kernel, mode = "code")
tccq_c_block(slice_sum_src)
```

The slice bind now lowers to an explicit `view1` node. In the current target,
that means the compiler can bind a pointer/length view and reduce over it
without allocating a copied slice first.

### Example: generic reducers and reducer-style idioms

```{r}
logic_reduce_kernel <- function(x, y) {
  declare(type(x = double(NA), y = double(NA)))
  all((x > 0) & (y > 0))
}

reduce_kernel <- function(x) {
  declare(type(x = double(NA)))
  Reduce(`+`, x)
}

compiled_logic_reduce <- tccq_compile(logic_reduce_kernel)
compiled_reduce <- tccq_compile(reduce_kernel)

list(
  logic_reduce = compiled_logic_reduce(c(1, 2, 3), c(4, 5, 6)),
  reduce_sum = compiled_reduce(c(1, 2, 3, 4))
)
```

This still does not claim full base-R `Reduce()` or `apply`-family coverage, but
it does mean the current compiler can translate a wider set of idiomatic
array-oriented reduction code than the earlier `sum()`-only milestone.

### Example: direct formal mutation is rejected

```{r}
direct_formal_mutation <- function(x, v) {
  declare(type(x = double(NA)), type(v = double()))
  x[1] <- v
  x
}

tryCatch(
  tccq_compile(direct_formal_mutation, mode = "code"),
  error = function(e) e$message
)
```

### Example: generate C source

```{r}
sum_src <- tccq_compile(fresh_sum_kernel, mode = "code")
tccq_c_block(sum_src)
```

The generated C contains one reduction loop and no intermediate vector
allocation for the reduction itself.

### Example: vector materialization kernel

```{r}
vec_src <- tccq_compile(fresh_vec_kernel, mode = "code")
tccq_c_block(vec_src)
```

This path allocates one output vector and fills it in one loop.

### Optional: compile through TinyCC

```{r}
if (requireNamespace("Rtinycc", quietly = TRUE)) {
  compiled_sum <- tccq_compile(fresh_sum_kernel)
  compiled_vec <- tccq_compile(fresh_vec_kernel)
  compiled_fallback <- tccq_compile(fallback_kernel, fallback = "auto")
  compiled_assign <- tccq_compile(assign_kernel)
  compiled_slice_sum <- tccq_compile(slice_sum_kernel)

  x <- as.double(seq(-2, 2, length.out = 10))
  y <- as.double(seq(1, 3, length.out = 10))

  list(
    compiled_sum = compiled_sum(x, y),
    compiled_vec_head = unname(compiled_vec(x, y)[1:4]),
    compiled_fallback = compiled_fallback(42),
    compiled_assign = unname(compiled_assign(c(1, 2, 3), 2L, 10)),
    compiled_slice_sum = compiled_slice_sum(c(1, 2, 3, 4), 2L, 4L)
  )
}
```

## Related projects and influences

`tccquickr` is not trying to be identical to any one upstream project, but its
current direction is easier to understand relative to a few concrete references.

### `quickr`

[`quickr`](https://github.com/t-kalinowski/quickr) is a useful comparison for a
declared R subset compiler. It demonstrates that a strict, typed subset of R
can be lowered aggressively and made fast.

### `anvil`

[`anvil`](https://github.com/r-xla/anvil) is a useful comparison for explicit
transformation architecture and backend thinking. It is broader than the goals
here, but it is a good reference point for separating frontend and backend
concerns.

### `SAC` and `sac2c`

[`SAC`](https://sac-home.org/about%3Asac) and the
[`sac2c`](https://gitlab.sac-home.org/sac-group/sac2c) compiler are the main
optimization inspiration for the current path here:

- explicit array/kernel IR
- fusion as an IR rewrite rather than a codegen accident
- allocation reduction and materialization discipline
- legality-driven optimization boundaries

## Status

This package is experimental.

The split from [`Rtinycc`](https://github.com/sounkou-bioinfo/Rtinycc) is about
separation of concerns:

- [`Rtinycc`](https://github.com/sounkou-bioinfo/Rtinycc) provides the TinyCC /
  FFI runtime layer
- `tccquickr` provides the compiler and transformation framework
- `tccq_*` is the active architecture for explicit IR, mutation barriers,
  slicing/indexing semantics, and backend-neutral compilation flow

## Development

`tccquickr` follows the same repo conventions as the parent package:

- `README.Rmd` as the source of the GitHub README
- `roxygen2` for documentation generation
- `tinytest` under `inst/tinytest`
- a simple package-local `Makefile` for build/check/test tasks
