---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", export_all = FALSE, helpers = FALSE, quiet = TRUE)
} else {
  library(tccquickr)
}
knitr::knit_engines$set(rtinycc = Rtinycc:::rtinycc_engine)

tccq_c_block <- function(code) {
  Rtinycc:::rtinycc_c_block(code)
}
```

# tccquickr: an R-to-C transformation framework

Experimental backend-neutral R-to-C transformation framework on top of
`Rtinycc`.

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/tccquickr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/tccquickr/actions/workflows/R-CMD-check.yaml)
[![tccquickr status badge](https://sounkou-bioinfo.r-universe.dev/tccquickr/badges/version)](https://sounkou-bioinfo.r-universe.dev/tccquickr/badges/version)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

## What this package is

`tccquickr` is an experimental R-to-C transformation framework for a small,
declared subset of R.

The package is centered on `tccq_compile()`, which takes
`declare(type(...))`-annotated R code, lowers it into explicit typed IR,
applies middle-end rewrites, emits C through the R C API, and then either:

- returns generated C source,
- compiles it in memory through TinyCC via `Rtinycc`, or
- compiles it as a shared library through `R CMD SHLIB`.

The public entry points are:

- `tccq_compile()`
- `tccq_backend_source()`
- `tccq_backend_tinycc()`
- `tccq_backend_shlib()`

Responsibility is split deliberately:

- `tccquickr` owns the frontend, IR, passes, C target, and backend-neutral
  orchestration
- [`Rtinycc`](https://github.com/sounkou-bioinfo/Rtinycc) owns the TinyCC
  toolchain, libtcc runtime, FFI pipeline, and low-level native runtime helpers

That means `tccquickr` is the transformation layer. TinyCC is one backend, not
the whole design.

## Installation

```{r eval = FALSE}
install.packages(
  "tccquickr",
  repos = c(
    "https://sounkou-bioinfo.r-universe.dev",
    "https://cloud.r-project.org"
  )
)
```

## Current scope

### Implemented today

- declared scalar, vector, and matrix inputs
- `double`, `integer`, and `logical` typing
- scalar and elementwise arithmetic
- comparison and logical vector expressions
- unary math calls such as `sin()`, `cos()`, `exp()`, `log()`, and `sqrt()`
- simple `for` loops over integer sequence ranges
- explicit kernel IR nodes such as `producer`, `materialize`, `fold`, and
  `scalar_kernel`
- statement/program IR for local bindings and writes
- scalar indexed reads `x[i]`
- vector gather reads `x[idx]`
- contiguous slices `x[lo:hi]`
- local indexed writes `y[i] <- v` and local range writes `y[lo:hi] <- v`
- scalar matrix reads `x[i, j]`
- matrix row and column extraction `x[i, ]` and `x[, j]`
- scalar, row, column, rectangle, and full-matrix local writes
- scalar vector fills such as `integer(n)`, `double(n)`, and `logical(n)`
- scalar `matrix(data, nrow, ncol)` construction
- fold-style reducers `sum`, `prod`, `min`, `max`, `mean`, `any`, and `all`
- arg reducers such as `which.max()`
- limited `Reduce(FUN, x)` lowering for recognized reducer surfaces
- explicit boundary nodes in `fallback = "auto"`
- source-only, TinyCC, and shared-library (`R CMD SHLIB`) backends
- compile-time backend/target capability validation
- hand-written and generated differential validation, including cross-backend
  checks

### Intentionally still limited

- full base-R `Reduce()` compatibility
- axis-wise reductions such as `rowSums()` / `colSums()`
- broader `apply`-family lowering
- richer reuse/allocation planning before C emission
- remaining broader indexing forms such as scatter/filter, vectorized matrix
  gather, and general matrix extraction
- larger harvested validation corpora from real array-oriented R code

## Quick tour

### Example 1: a reduction kernel

Start with a kernel that reduces a vector expression to one scalar.

```{r}
sum_kernel <- function(x, y) {
  declare(type(x = double(NA), y = double(NA)))
  sum((sin(x) + y) * y)
}
```

### Inspect the IR

```{r}
sum_module <- tccq_compile(sum_kernel, mode = "ir")
sum_module
```

This path lowers to an explicit `fold` over a `producer`.

### Generate C

<details>
<summary>Show generated C for <code>sum_kernel</code></summary>

```{r}
sum_src <- tccq_compile(sum_kernel, mode = "code")
tccq_c_block(sum_src)
```

</details>

### Compile and run it

```{r}
compiled_sum_kernel <- tccq_compile(sum_kernel)
x <- as.double(seq(-2, 2, length.out = 10))
y <- as.double(seq(1, 3, length.out = 10))
compiled_sum_kernel(x, y)
```

### Example 2: a vector-return kernel

Now take a kernel that produces a whole vector result.

```{r}
vec_kernel <- function(x, y) {
  declare(type(x = double(NA), y = double(NA)))
  sin(x) + y * y
}
```

### Inspect the IR

```{r}
vec_module <- tccq_compile(vec_kernel, mode = "ir")
vec_module
```

This path lowers to `materialize(producer)`.

### Generate C

<details>
<summary>Show generated C for <code>vec_kernel</code></summary>

```{r}
vec_src <- tccq_compile(vec_kernel, mode = "code")
tccq_c_block(vec_src)
```

</details>

### Compile and run it

```{r}
compiled_vec_kernel <- tccq_compile(vec_kernel)
unname(compiled_vec_kernel(x, y)[1:4])
```

### Example 3: a Viterbi-style dynamic program

This example follows the shape of the classic `quickr` Viterbi example, but it
keeps the explicit declared-subset contract in `tccquickr`. The important compiler
features here are general, not example-specific: braced declarations,
matrix row/column views, matrix writes, local vector fills, nested `for` loops,
`max()`, and `which.max()`.

```{r}
viterbi_r <- function(
  observations,
  states,
  initial_probs,
  transition_probs,
  emission_probs
) {
  num_states <- length(states)
  num_steps <- length(observations)

  trellis <- matrix(0, nrow = length(states), ncol = length(observations))
  backpointer <- matrix(0L, nrow = length(states), ncol = length(observations))

  trellis[, 1] <- initial_probs * emission_probs[, observations[1]]

  for (step in 2:num_steps) {
    for (current_state in 1:num_states) {
      probabilities <- trellis[, step - 1] * transition_probs[, current_state]
      trellis[current_state, step] <- max(probabilities) *
        emission_probs[current_state, observations[step]]
      backpointer[current_state, step] <- which.max(probabilities)
    }
  }

  path <- integer(length(observations))
  path[num_steps] <- which.max(trellis[, num_steps])
  for (step in seq((num_steps - 1), 1)) {
    path[step] <- backpointer[path[step + 1], step + 1]
  }

  states[path]
}

viterbi_tccq <- function(
  observations,
  states,
  initial_probs,
  transition_probs,
  emission_probs
) {
  declare({
    type(observations = integer(num_steps))
    type(states = integer(num_states))
    type(initial_probs = double(num_states))
    type(transition_probs = double(num_states, num_states))
    type(emission_probs = double(num_states, num_obs))
  })

  num_states <- length(states)
  num_steps <- length(observations)

  trellis <- matrix(0, nrow = length(states), ncol = length(observations))
  backpointer <- matrix(0L, nrow = length(states), ncol = length(observations))

  trellis[, 1] <- initial_probs * emission_probs[, observations[1]]

  for (step in 2:num_steps) {
    for (current_state in 1:num_states) {
      probabilities <- trellis[, step - 1] * transition_probs[, current_state]
      trellis[current_state, step] <- max(probabilities) *
        emission_probs[current_state, observations[step]]
      backpointer[current_state, step] <- which.max(probabilities)
    }
  }

  path <- integer(length(observations))
  path[num_steps] <- which.max(trellis[, num_steps])
  for (step in seq((num_steps - 1), 1)) {
    path[step] <- backpointer[path[step + 1], step + 1]
  }

  states[path]
}
```

Compile once, then call the compiled closure like a normal R function.

```{r}
set.seed(42)
num_steps <- 50L
num_states <- 6L
num_obs <- 20L

observations <- sample.int(num_obs, num_steps, replace = TRUE)
states <- seq_len(num_states)
initial_probs <- runif(num_states)
initial_probs <- initial_probs / sum(initial_probs)
transition_probs <- matrix(runif(num_states * num_states), nrow = num_states)
transition_probs <- transition_probs / rowSums(transition_probs)
emission_probs <- matrix(runif(num_states * num_obs), nrow = num_states)
emission_probs <- emission_probs / rowSums(emission_probs)

compiled_viterbi <- tccq_compile(viterbi_tccq)
identical(
  compiled_viterbi(
    observations, states, initial_probs, transition_probs, emission_probs
  ),
  viterbi_r(
    observations, states, initial_probs, transition_probs, emission_probs
  )
)
```

### Viterbi runtime benchmark

This benchmark intentionally excludes compilation time. It compares direct R
evaluation with repeated calls to the already compiled `tccquickr` closure. It
uses batched `system.time()` so the README benchmark has no optional runtime
dependency.

```{r}
readme_bench <- function(expr, batches = 30L, batch_size = 100L) {
  expr <- substitute(expr)
  env <- parent.frame()
  timings <- vapply(seq_len(batches), function(batch) {
    unname(system.time({
      for (i in seq_len(batch_size)) {
        eval(expr, env)
      }
    })[["elapsed"]]) / batch_size
  }, numeric(1))
  median(timings)
}

viterbi_elapsed <- c(
  R = readme_bench(
    viterbi_r(
      observations, states, initial_probs, transition_probs, emission_probs
    )
  ),
  tccquickr = readme_bench(
    compiled_viterbi(
      observations, states, initial_probs, transition_probs, emission_probs
    )
  )
)

data.frame(
  expression = names(viterbi_elapsed),
  median_us = round(viterbi_elapsed * 1e6, 1),
  itr_sec = round(1 / viterbi_elapsed)
)
```

## Backend selection

`tccq_compile()` can target different C backends explicitly.

```{r eval = FALSE}
tccq_compile(f, backend = tccq_backend_source())
tccq_compile(f, backend = tccq_backend_tinycc())
tccq_compile(f, backend = tccq_backend_shlib())
```

The roles are:

- `tccq_backend_source()` returns emitted C and does not compile it
- `tccq_backend_tinycc()` compiles emitted C in memory through `Rtinycc`
- `tccq_backend_shlib()` compiles emitted C through `R CMD SHLIB` and loads it
  through `.Call()`

The shared-library route is in the same general deployment space as
[`callme`](https://github.com/coolbutuseless/callme): a C-only compile/load
workflow around shared objects and `.Call()` entry points.

Before compilation, `tccq_compile()` validates backend capabilities against the
selected target, compile context, and any explicit boundary APIs.

## Current semantic model

The current subset is small on purpose. The key rules are explicit and are meant
to live in IR and middle-end plans rather than appearing accidentally in the C
emitter.

### Bindings, reads, slices, and writes

- `a <- expr` is a local binding, not mutation
- `x[i]` is a checked 1-based indexed read
- `x[lo:hi]` is a contiguous slice/view expression
- `a[i] <- v` and `a[lo:hi] <- v` are mutation barriers
- direct mutation of formals such as `x[i] <- v` is rejected for now

### Example: local bind, then indexed write

```{r}
assign_kernel <- function(x, i, v) {
  declare(type(x = double(NA)), type(i = integer()), type(v = double()))
  y <- x
  y[i] <- v
  y
}
```

Compile and run it:

```{r}
compiled_assign_kernel <- tccq_compile(assign_kernel)
unname(compiled_assign_kernel(c(1, 2, 3), 2L, 10))
```

Show the generated C:

<details>
<summary>Show generated C for <code>assign_kernel</code></summary>

```{r}
assign_src <- tccq_compile(assign_kernel, mode = "code")
tccq_c_block(assign_src)
```

</details>

Here `y <- x` starts as an alias to the formal input. The first indexed write
forces materialization into owned local storage.

### Example: slice, then reduce

```{r}
slice_sum_kernel <- function(x, lo, hi) {
  declare(type(x = double(NA)), type(lo = integer()), type(hi = integer()))
  y <- x[lo:hi]
  sum(y)
}
```

Compile and run it:

```{r}
compiled_slice_sum_kernel <- tccq_compile(slice_sum_kernel)
compiled_slice_sum_kernel(c(1, 2, 3, 4), 2L, 4L)
```

Show the generated C:

<details>
<summary>Show generated C for <code>slice_sum_kernel</code></summary>

```{r}
slice_sum_src <- tccq_compile(slice_sum_kernel, mode = "code")
tccq_c_block(slice_sum_src)
```

</details>

This slice bind lowers to an explicit `view1` node. In the current target, that
lets the compiler reduce over the slice extent without first copying the slice
into a fresh vector.

### Reducers and reducer-style idioms

Reducers are lowered through explicit fold machinery, not through one-off
`sum()`-only code generation branches.

Supported direct reducers include:

- `sum`, `prod`, `min`, `max`, `mean`
- `any`, `all`

Recognized `Reduce(FUN, x)` surfaces currently include reducer-equivalent forms
such as:

- `Reduce(`+`, x)`
- `Reduce(`*`, x)`
- `Reduce(`&`, x)`
- `Reduce(`|`, x)`
- `Reduce(min, x)`
- `Reduce(max, x)`

```{r}
logic_reduce_kernel <- function(x, y) {
  declare(type(x = double(NA), y = double(NA)))
  all((x > 0) & (y > 0))
}

compiled_logic_reduce_kernel <- tccq_compile(logic_reduce_kernel)
compiled_logic_reduce_kernel(c(1, 2, 3), c(4, 5, 6))
```

```{r}
reduce_kernel <- function(x) {
  declare(type(x = double(NA)))
  Reduce(`+`, x)
}

compiled_reduce_kernel <- tccq_compile(reduce_kernel)
compiled_reduce_kernel(c(1, 2, 3, 4))
```

This is still intentionally narrower than full base-R `Reduce()` semantics.

### Unsupported calls and explicit boundaries

Unsupported calls are not supposed to disappear silently into target-specific
special cases. The default `fallback = "hard"` rejects unsupported calls. With
explicit `fallback = "auto"`, they lower to inspectable boundary nodes.

```{r}
fallback_kernel <- function(x) {
  declare(type(x = double()))
  floor(x)
}

fallback_ir <- tccq_compile(fallback_kernel, mode = "ir", fallback = "auto")
fallback_ir$ir$result$tag
fallback_ir$ir$result$api
fallback_ir$boundary_diagnostics[[1]]$original_call
```

In default hard-fallback mode, the same unsupported call is rejected.

```{r}
tryCatch(
  tccq_compile(fallback_kernel, mode = "ir"),
  error = function(e) e$message
)
```

### Direct formal mutation is rejected

```{r}
direct_formal_mutation <- function(x, v) {
  declare(type(x = double(NA)), type(v = double()))
  x[1] <- v
  x
}

tryCatch(
  tccq_compile(direct_formal_mutation, mode = "code"),
  error = function(e) e$message
)
```

## Architecture

`tccquickr` is organized as an explicit transformation pipeline rather than as a
single direct “R expression to C string” step.

### Pipeline

- **frontend**: parse `declare(type(...))` and lower a small R subset into typed
  IR
- **middle-end**: validate IR, infer effects, normalize access paths, derive
  explicit shape/domain facts, lower reducers, kernelize, fuse, handle explicit
  boundaries, and derive conservative storage/protection plans
- **target**: emit C through the R C API
- **backend**: either return source or compile the emitted C through TinyCC or
  `R CMD SHLIB`

### Kernel IR concepts

`producer`
: an element computation over an index domain

`materialize(producer)`
: allocate an output vector and write producer elements into it

`fold(op, domain, elem)`
: reduce a producer-like element expression over a loop domain

These are the concepts that make fusion and materialization decisions explicit
in IR instead of hiding them inside the C emitter.

### Fusion is an explicit rewrite

Before fusion, the middle-end now makes access chains and shape/domain facts
explicit, then `kernelize` represents the reduction as a fold over a
materialized producer. The fusion pass rewrites that to a fold over the
producer directly.

```{r}
mod0 <- tccquickr:::tccq_frontend(sum_kernel)
mod0 <- tccquickr:::tccq_lower_module(mod0)

before_fusion <- tccquickr:::tccq_run_passes(
  mod0,
  passes = list(
    tccquickr:::tccq_pass_validate_ir(),
    tccquickr:::tccq_pass_effects(),
    tccquickr:::tccq_pass_index_normalize(),
    tccquickr:::tccq_pass_shape_domains(),
    tccquickr:::tccq_pass_kernelize()
  )
)

after_fusion <- tccquickr:::tccq_run_passes(
  mod0,
  passes = list(
    tccquickr:::tccq_pass_validate_ir(),
    tccquickr:::tccq_pass_effects(),
    tccquickr:::tccq_pass_index_normalize(),
    tccquickr:::tccq_pass_shape_domains(),
    tccquickr:::tccq_pass_kernelize(),
    tccquickr:::tccq_pass_fusion()
  )
)

before_fusion
after_fusion
```

That rewrite system is still small, but the important architectural point is
that fusion lives in the middle-end rather than being discovered accidentally
while printing C.

## Validation

Validation is part of the architecture, not an afterthought.

The current package combines:

- hand-written runtime tests for reducers, slicing, views, boundaries, storage
  behavior, and backends
- generated differential tests that construct typed R functions
  programmatically and compare compiled behavior against direct R evaluation
- cross-backend checks between TinyCC and `R CMD SHLIB`

The goal is not to claim formal proof. The goal is to keep the transformation
pipeline honest with executable witness tests across IR, emitted C, and runtime
behavior.

## Related projects and influences

These are reference points, not templates that `tccquickr` is trying to copy
exactly.

- [`quickr`](https://github.com/t-kalinowski/quickr): useful comparison for a
  declared R subset compiler
- [`anvil`](https://github.com/r-xla/anvl): useful comparison for explicit
  transformation architecture and backend thinking
- [`callme`](https://github.com/coolbutuseless/callme): useful reference for a
  shared-library compile/load workflow around `.Call()`
- [`SAC`](https://sac-home.org/) and
  [`sac2c`](https://gitlab.sac-home.org/sac-group/sac2c): main inspiration for
  explicit array IR, fusion, and allocation/materialization discipline

## Development

`tccquickr` follows the same basic repo conventions as the parent package:

- `README.Rmd` is the source of `README.md`
- `roxygen2` generates documentation
- `tinytest` lives under `inst/tinytest`
- a simple package-local `Makefile` drives common build/check/test tasks
