---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# tccquickr

`tccquickr` compiles a declared subset of R into typed IR and then into C,
Fortran, and TinyCC-JIT kernels. Programs enter through `declare(type(...))`,
become S7 value graphs with symbolic shapes and effects, and lower into one
backend-neutral loop nest — a SAC-style with-loop — before any target syntax
exists. Unsupported R is a classed diagnostic, never an implicit fallback.

Every chunk in this README executes when the document is knit, so the claims
below are checked against the installed package, not maintained by hand.

## Install

```r
install.packages(
  "tccquickr",
  repos = c("https://sounkou-bioinfo.r-universe.dev", getOption("repos"))
)
# or
remotes::install_github("sounkou-bioinfo/tccquickr")
```

## From declared R to a running kernel

A matrix-vector product, declared with symbolic dimensions:

```{r declare}
library(tccquickr)

matrix_vector <- function(x, w) {
  declare(type(x = double(n, p), w = double(p)))
  x %*% w
}
```

The analysis lowers this into a typed loop nest — one `map` axis over `n`, one
`reduce` axis over `p` folding with `sum` — and the C backend prints that nest:

```{r generated-c}
program <- tccq_analyze(matrix_vector, strict = TRUE)@value
c_plan <- tccq_plan_backend(
  program,
  tccq_c_backend(),
  tccq_backend_context(mode = "source", target = "c")
)
cat(c_plan@value@products@attrs$source)
```

The extent parameters come from the declared dimension symbols, so the
generated boundary knows that `ncol(x)` and `length(w)` are the same `p` and
checks it. TinyCC compiles the same source in-process and the callable agrees
with R:

```{r run-jit}
jit_plan <- tccq_plan_backend(
  program,
  tccq_rtinycc_backend(),
  tccq_backend_context(mode = "jit", target = "c")
)
matvec <- jit_plan@value@products@attrs$callable

set.seed(1)
x <- matrix(rnorm(12), nrow = 3)
w <- rnorm(4)
all.equal(matvec(x, w), drop(x %*% w))
```

Stencils work the same way: affine slice bounds become typed affine extents
(`n - 2` is a dimension fact, not a printed string), and the loop nest shifts
accesses instead of materializing slices:

```{r stencil}
tiled_stencil_1d <- function(x) {
  declare(type(x = double(n)))
  x[1:(n - 2L)] + x[2:(n - 1L)] + x[3:n]
}

stencil_jit <- tccq_plan_backend(
  tccq_analyze(tiled_stencil_1d, strict = TRUE)@value,
  tccq_rtinycc_backend(),
  tccq_backend_context(mode = "jit", target = "c")
)
stencil <- stencil_jit@value@products@attrs$callable
v <- c(1, 2, 4, 8, 16, 32, 64)
all.equal(stencil(v), v[1:5] + v[2:6] + v[3:7])
```

## The declared subset today

- **Types**: `double` formals of any rank with symbolic or constant
  dimensions. Other declared base types analyze but do not reach the source
  backends yet.
- **Elementwise**: `+`, `-`, `*`, `/`, `^`, `sqrt`, `exp` with scalar
  broadcast over any rank; registries can add rendered operations without
  touching the printers.
- **Reductions**: `sum` over full rank-N domains; `colSums` and `rowSums` as
  per-axis reductions; custom reducers via `TccqReductionSpec`.
- **Contractions**: `%*%` for matrix-vector and matrix-matrix products with a
  typed inner-dimension rule (`TccqContractionSpec`).
- **Slices**: rank-1 `x[a:b]` with bounds affine in declared dimension
  symbols.
- **Bindings**: single-assignment locals.

Reductions and contractions must produce the program result; anything needing
more than one loop nest is a structured `lowering.unsupported_composition`
diagnostic until multi-nest programs are modeled. Compilation succeeds when at
least one backend produces a working plan; a backend that cannot lower a
program reports typed feasibility diagnostics without vetoing the suite.

## Apotheosis suite

The targets are explicit probes, each exercising one compiler idea, plus
composite programs that force the ideas together. The table below is computed
at knit time by running every probe through `tccq_compile()`; a probe either
compiles through the core backends or names the diagnostic where it stops.

```{r probes}
probes <- list(
  map_chain = function(x, y) {
    declare(type(x = double(n), y = double(n)))
    exp(sqrt(x) + y)
  },
  map_reduce = function(x, y) {
    declare(type(x = double(n), y = double(n)))
    sum(exp(x) * y)
  },
  matrix_reduce = function(x, y) {
    declare(type(x = double(n, p), y = double(n, p)))
    sum(exp(x) * y)
  },
  matrix_map = function(x, y) {
    declare(type(x = double(n, p), y = double(n, p)))
    sqrt(x) + y
  },
  column_sums = function(x) {
    declare(type(x = double(n, p)))
    colSums(x)
  },
  matrix_vector = function(x, w) {
    declare(type(x = double(n, p), w = double(p)))
    x %*% w
  },
  matrix_multiply = function(x, y) {
    declare(type(x = double(n, p), y = double(p, q)))
    x %*% y
  },
  tiled_stencil_1d = function(x) {
    declare(type(x = double(n)))
    x[1:(n - 2L)] + x[2:(n - 1L)] + x[3:n]
  },
  raw_buffer_roundtrip = function(bytes, scratch) {
    declare(type(bytes = raw(n), scratch = buffer(n)))
    bytes
  },
  control_flow_probe = function(x, flag) {
    declare(type(x = double(n), flag = logical()))
    out <- 0
    i <- 1L
    repeat {
      if (i > n) {
        break
      }
      out <- out + ifelse(flag, x[i], -x[i])
      i <- i + 1L
    }
    switch(if (flag) "sum" else "neg", sum = out, neg = -out)
  },
  apply_reduce_probe = function(x) {
    declare(type(x = double(n, p)))
    Reduce(`+`, lapply(seq_len(p), function(j) sum(x[, j])))
  },
  logistic_gradient = function(x, y, w, lambda) {
    declare(type(
      x = double(n, p),
      y = double(n),
      w = double(p),
      lambda = double()
    ))
    mu <- colMeans(x)
    sigma <- sqrt(colSums((x - mu)^2) / (n - 1L))
    z <- (x - mu) / sigma
    eta <- z %*% w
    prob <- 1 / (1 + exp(-eta))
    grad <- crossprod(z, prob - y) / n + lambda * w
    w - 0.01 * grad
  },
  viterbi_decode = function(init, trans, emit) {
    declare(type(
      init = double(k),
      trans = double(k, k),
      emit = double(t, k)
    ))
    score <- init + emit[1L, ]
    back <- matrix(0L, nrow = t, ncol = k)
    for (step in 2:t) {
      next_score <- score
      for (state in 1:k) {
        candidate <- score + trans[, state]
        best <- which.max(candidate)
        next_score[state] <- candidate[best] + emit[step, state]
        back[step, state] <- best
      }
      score <- next_score
    }
    list(score = score, back = back)
  }
)
```

```{r apotheosis-table, results = "asis"}
probe_status <- function(fn) {
  compiled <- tccq_compile(fn, strict = FALSE)
  if (compiled@success) {
    return("compiles through C, Fortran, and TinyCC JIT")
  }
  diagnostic <- compiled@diagnostics[[1L]]
  sprintf("`%s`", diagnostic@code)
}
knitr::kable(data.frame(
  probe = names(probes),
  status = vapply(probes, probe_status, character(1)),
  row.names = NULL
))
```

The failing rows are the roadmap: every one must move deeper through the same
typed IR — general indexing and access regions, structured control flow,
`raw`/`buffer` bridges, apply-family folds, and finally multi-nest composition
for the composite targets. Failures must stay specific enough that the next
typed concept to add is obvious.

## Design

The compiler core is typed all the way down: S7 schemas for values, shapes,
effects, regions, and plans; `s7contract` interfaces between phases; classed
diagnostics instead of error strings; and one `TccqLoopNest` iteration
abstraction consumed by every source backend. The full phase-by-phase story —
call semantics, operation registries, lowering, loop nests, bridges, backend
planning, fusion — lives in
[docs/architecture.md](docs/architecture.md), with the root direction in
[docs/root.md](docs/root.md) and repo rules in [AGENTS.md](AGENTS.md).
