---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", export_all = FALSE, helpers = FALSE, quiet = TRUE)
} else {
  library(tccquickr)
}
```

# tccquickr

Experimental backend-neutral R code transformation framework on top of
`Rtinycc`.

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/tccquickr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/tccquickr/actions/workflows/R-CMD-check.yaml)
[![tccquickr status badge](https://sounkou-bioinfo.r-universe.dev/tccquickr/badges/version)](https://sounkou-bioinfo.r-universe.dev/tccquickr)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

## Abstract

`tccquickr` is an experimental compiler and transformation framework for a
small, declared R subset.

The current design direction is centered on the fresh `tccq2_*` path:

- frontend parsing and typed lowering for `declare(type(...))`-annotated R
- explicit middle-end IR for producers, materialization, folds, statements, and
  legality barriers
- C emission through the R C API
- swappable backends, with source-only output and TinyCC via
  [`Rtinycc`](https://github.com/sounkou-bioinfo/Rtinycc)

The package depends on [`Rtinycc`](https://github.com/sounkou-bioinfo/Rtinycc)
for the underlying TinyCC toolchain, libtcc runtime, FFI compilation pipeline,
and pointer/runtime support.

Conceptually, the project sits near several adjacent efforts:

- [`quickr`](https://github.com/t-kalinowski/quickr) as a useful R subset
  compiler comparison
- [`anvil`](https://github.com/r-xla/anvil) as a broader transformation and
  backend framework comparison
- [`SAC` / `sac2c`](https://sac-home.org/about%3Asac) as the main inspiration
  for explicit array IR, fusion, and allocation reduction

## Scope

`tccquickr` is where the transformation framework lives.

The main moving parts are:

- `tccq2_compile()`
- typed frontend parsing and lowering
- middle-end kernel IR and rewrite passes
- C + R C API target emission
- tinytest coverage for generated code and compiled behavior

[`Rtinycc`](https://github.com/sounkou-bioinfo/Rtinycc) remains responsible for:

- bundled TinyCC build and installation
- `tcc_ffi()`, `tcc_bind()`, and `tcc_compile()`
- low-level state and CLI helpers
- pointer, struct, callback, and declarative FFI support

## Installation

```{r eval = FALSE}
install.packages(
  "tccquickr",
  repos = c(
    "https://sounkou-bioinfo.r-universe.dev",
    "https://cloud.r-project.org"
  )
)
```

## Fresh `tccq2` scaffold

The fresh path is deliberately small in the first milestones.

Milestone 1 / 2 supports:

- declared scalar and vector inputs
- `double`, `integer`, and `logical` typing
- scalar arithmetic
- vector elementwise arithmetic
- unary math calls such as `sin()`, `cos()`, `exp()`, `log()`, and `sqrt()`
- `sum(...)` as a fused reduction
- explicit kernel IR nodes for `producer`, `materialize`, and `fold`
- statement blocks with local bindings
- scalar indexed reads `x[i]`
- contiguous range slices `x[lo:hi]`
- local indexed writes `y[i] <- v` and local range writes `y[lo:hi] <- v`

The intended split is:

- frontend: parse `declare(type(...))` and lower a small R subset
- middle-end: validate, kernelize, and later perform explicit fusion and
  allocation planning
- target: emit C using the R C API
- backend: either return source or compile through TinyCC via `Rtinycc`

### Kernel IR concepts

`producer`
: an element computation over an index domain

`materialize(producer)`
: allocate an output vector and write producer elements into it

`fold(op, domain, elem)`
: reduce a producer-like element expression over a loop domain

These are the concepts that let the package move from “fusion hidden in code
emission” toward “fusion represented in IR”.

### Assignment and slicing rules

The fresh compiler treats these forms differently:

- `a <- expr` is a local binding, not mutation
- `x[i]` is an indexed read with checked 1-based indexing
- `x[lo:hi]` is a contiguous slice expression
- `a[i] <- v` and `a[lo:hi] <- v` are mutation barriers

For the current milestone, indexed writes are only allowed into local vector
bindings. Direct formal mutation such as `x[i] <- v` is rejected.

### Example: inspect fresh IR

```{r}
fresh_sum_kernel <- function(x, y) {
  declare(type(x = double(NA), y = double(NA)))
  sum((sin(x) + y) * y)
}

fresh_vec_kernel <- function(x, y) {
  declare(type(x = double(NA), y = double(NA)))
  sin(x) + y * y
}

sum_module <- tccq2_compile(fresh_sum_kernel, mode = "ir")
vec_module <- tccq2_compile(fresh_vec_kernel, mode = "ir")

list(
  sum_kernel_tag = sum_module$kernel$tag,
  sum_domain_tag = sum_module$kernel$domain$tag,
  sum_elem_tag = sum_module$kernel$elem$tag,
  vec_kernel_tag = vec_module$kernel$tag,
  vec_producer_tag = vec_module$kernel$producer$tag
)
```

The reduction path is represented as a `fold` over an explicit `producer`.
The vector return path is represented as `materialize(producer)`.

### Example: explicit fusion rewrite

Before fusion, `kernelize` represents the reduction as a fold over a
materialized producer. The fusion pass rewrites that to a fold over the
producer directly.

```{r}
mod0 <- tccquickr:::tccq2_frontend(fresh_sum_kernel)
mod0 <- tccquickr:::tccq2_lower_module(mod0)

before_fusion <- tccquickr:::tccq2_run_passes(
  mod0,
  passes = list(
    tccquickr:::tccq2_pass_validate_ir(),
    tccquickr:::tccq2_pass_effects(),
    tccquickr:::tccq2_pass_kernelize()
  )
)

after_fusion <- tccquickr:::tccq2_run_passes(
  mod0,
  passes = list(
    tccquickr:::tccq2_pass_validate_ir(),
    tccquickr:::tccq2_pass_effects(),
    tccquickr:::tccq2_pass_kernelize(),
    tccquickr:::tccq2_pass_fusion()
  )
)

list(
  before = before_fusion$kernel$elem$tag,
  after = after_fusion$kernel$elem$tag
)
```

This is still a small rewrite system, but the important point is that fusion is
now expressed as a middle-end transformation rather than being buried entirely
inside C string emission.

### Example: legality barriers

Boundary nodes are explicit legality barriers. The fresh compiler does not yet
fallback through them; it refuses to compile them.

```{r}
boundary_ir <- tccquickr:::tccq2_ir_boundary(
  kind = "rf_call",
  reason = "unsupported R call",
  input = tccquickr:::tccq2_ir_var("x", tccquickr:::tccq2_type_vector("double")),
  type = tccquickr:::tccq2_type_vector("double")
)

boundary_module <- tccquickr:::tccq2_module(
  entry = "tccq2_entry",
  formal_names = c("x"),
  types = list(x = tccquickr:::tccq2_type_vector("double")),
  expr = quote(x),
  ir = boundary_ir
)

tryCatch(
  tccquickr:::tccq2_run_passes(boundary_module),
  error = function(e) e$message
)
```

### Example: assignments and indexed writes

```{r}
assign_kernel <- function(x, i, v) {
  declare(type(x = double(NA)), type(i = integer()), type(v = double()))
  y <- x
  y[i] <- v
  y
}

assign_src <- tccq2_compile(assign_kernel, mode = "code")
cat(assign_src)
```

The local `y <- x` binding is materialized into owned local storage before the
indexed write. That keeps direct writes to caller-owned formal vectors out of
this milestone.

### Example: slicing and reduction

```{r}
slice_sum_kernel <- function(x, lo, hi) {
  declare(type(x = double(NA)), type(lo = integer()), type(hi = integer()))
  sum(x[lo:hi])
}

slice_sum_src <- tccq2_compile(slice_sum_kernel, mode = "code")
cat(slice_sum_src)
```

### Example: direct formal mutation is rejected

```{r}
direct_formal_mutation <- function(x, v) {
  declare(type(x = double(NA)), type(v = double()))
  x[1] <- v
  x
}

tryCatch(
  tccq2_compile(direct_formal_mutation, mode = "code"),
  error = function(e) e$message
)
```

### Example: generate C source

```{r}
sum_src <- tccq2_compile(fresh_sum_kernel, mode = "code")
cat(sum_src)
```

The generated C contains one reduction loop and no intermediate vector
allocation for the reduction itself.

### Example: vector materialization kernel

```{r}
vec_src <- tccq2_compile(fresh_vec_kernel, mode = "code")
cat(vec_src)
```

This path allocates one output vector and fills it in one loop.

### Optional: compile through TinyCC

```{r}
if (requireNamespace("Rtinycc", quietly = TRUE)) {
  compiled_sum <- tccq2_compile(fresh_sum_kernel)
  compiled_vec <- tccq2_compile(fresh_vec_kernel)

  x <- as.double(seq(-2, 2, length.out = 10))
  y <- as.double(seq(1, 3, length.out = 10))

  list(
    compiled_sum = compiled_sum(x, y),
    compiled_vec_head = unname(compiled_vec(x, y)[1:4])
  )
}
```

## Related projects and influences

`tccquickr` is not trying to be identical to any one upstream project, but its
current direction is easier to understand relative to a few concrete references.

### `quickr`

[`quickr`](https://github.com/t-kalinowski/quickr) is a useful comparison for a
declared R subset compiler. It demonstrates that a strict, typed subset of R
can be lowered aggressively and made fast.

### `anvil`

[`anvil`](https://github.com/r-xla/anvil) is a useful comparison for explicit
transformation architecture and backend thinking. It is broader than the goals
here, but it is a good reference point for separating frontend and backend
concerns.

### `SAC` and `sac2c`

[`SAC`](https://sac-home.org/about%3Asac) and the
[`sac2c`](https://gitlab.sac-home.org/sac-group/sac2c) compiler are the main
optimization inspiration for the fresh path here:

- explicit array/kernel IR
- fusion as an IR rewrite rather than a codegen accident
- allocation reduction and materialization discipline
- legality-driven optimization boundaries

## Status

This package is experimental.

The split from [`Rtinycc`](https://github.com/sounkou-bioinfo/Rtinycc) is about
separation of concerns:

- [`Rtinycc`](https://github.com/sounkou-bioinfo/Rtinycc) provides the TinyCC /
  FFI runtime layer
- `tccquickr` provides the compiler and transformation framework
- `tccq2_*` is the active architecture for explicit IR, mutation barriers,
  slicing/indexing semantics, and backend-neutral compilation flow

## Development

`tccquickr` follows the same repo conventions as the parent package:

- `README.Rmd` as the source of the GitHub README
- `roxygen2` for documentation generation
- `tinytest` under `inst/tinytest`
- a simple package-local `Makefile` for build/check/test tasks
