---
output: github_document
---

# tccquickr

`tccquickr` is being hard-reset as a typed R transformation core.

The new root is:

- declared R programs, not source editing;
- S7 schemas for compiler values;
- `s7contract` interfaces for internal compiler protocols;
- classed diagnostic values instead of stringly errors;
- shape-aware analysis before any C emission;
- no fake backward compatibility while the semantic core is rebuilt.

The current targets are not hidden package functions. They are explicit
apotheosis examples: small probes that exercise one compiler idea at a time,
and larger programs that force those ideas to compose. They cannot compile
today. Making these programs pass through a typed IR is the first serious
milestone.

## Minimal probes

These examples are intentionally small. Each should move from structured
failure to typed IR before we add more surface area.

```{r eval = FALSE}
map_chain <- function(x, y) {
  declare(type(x = double(n), y = double(n)))
  exp(sqrt(x) + y)
}

map_reduce <- function(x, y) {
  declare(type(x = double(n), y = double(n)))
  sum(exp(x) * y)
}

matrix_vector <- function(x, w) {
  declare(type(x = double(n, p), w = double(p)))
  x %*% w
}

raw_buffer_roundtrip <- function(bytes, scratch) {
  declare(type(bytes = raw(n), scratch = buffer(n)))
  bytes
}

tiled_stencil_1d <- function(x) {
  declare(type(x = double(n)))
  x[1:(n - 2L)] + x[2:(n - 1L)] + x[3:n]
}
```

## Composite targets

The larger examples force shapes, domains, layouts, reductions, regions,
operation implementations, bridges, and fusion planning to interact.

```{r eval = FALSE}
logistic_gradient <- function(x, y, w, lambda) {
  declare(type(
    x = double(n, p),
    y = double(n),
    w = double(p),
    lambda = double()
  ))

  mu <- colMeans(x)
  sigma <- sqrt(colSums((x - mu)^2) / (n - 1L))
  z <- (x - mu) / sigma
  eta <- z %*% w
  prob <- 1 / (1 + exp(-eta))
  grad <- crossprod(z, prob - y) / n + lambda * w
  w - 0.01 * grad
}

viterbi_decode <- function(init, trans, emit) {
  declare(type(
    init = double(k),
    trans = double(k, k),
    emit = double(t, k)
  ))

  score <- init + emit[1L, ]
  back <- matrix(0L, nrow = t, ncol = k)
  for (step in 2:t) {
    next_score <- score
    for (state in 1:k) {
      candidate <- score + trans[, state]
      best <- which.max(candidate)
      next_score[state] <- candidate[best] + emit[step, state]
      back[step, state] <- best
    }
    score <- next_score
  }
  list(score = score, back = back)
}

tccq_analyze(logistic_gradient)
tccq_compile(logistic_gradient)

tccq_analyze(viterbi_decode)
tccq_compile(viterbi_decode)
```

The expected state is structured failure: unsupported operations should be
reported as classed diagnostics until the frontend, typed IR, region planner,
operation implementations, bridges, and backend feasibility checks can account
for them honestly.
