---
title: "Getting started with tccquickr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with tccquickr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tccquickr)
```

`tccquickr` compiles a small, **declared** subset of R to C. You annotate a
function's argument types with `declare(type(...))`, and `tccq_compile()` returns
a compiled closure you call like any R function. This vignette is a tour; for the
exact accepted subset and the boundary/optimization model see
`vignette("the-r-subset")`.

## A reduction kernel

A kernel that reduces a vector expression to one scalar:

```{r}
sum_kernel <- function(x, y) {
  declare(type(x = double(NA), y = double(NA)))
  sum((sin(x) + y) * y)
}

compiled_sum <- tccq_compile(sum_kernel)
x <- as.double(seq(-2, 2, length.out = 10))
y <- as.double(seq(1, 3, length.out = 10))
compiled_sum(x, y)
```

You can stop before the backend and inspect the intermediate representation or
the emitted C:

```{r}
tccq_compile(sum_kernel, mode = "ir")
```

```{r}
cat(tccq_compile(sum_kernel, mode = "code"))
```

This path lowers to an explicit `fold` over a `producer` — the reduction and its
element expression are represented in the IR, not discovered while printing C.

## A vector-return kernel

A kernel that produces a whole vector lowers to `materialize(producer)`:

```{r}
vec_kernel <- function(x, y) {
  declare(type(x = double(NA), y = double(NA)))
  sin(x) + y * y
}

compiled_vec <- tccq_compile(vec_kernel)
unname(compiled_vec(x, y)[1:4])
```

## Choosing a backend

`tccq_compile()` targets different C backends explicitly:

```{r, eval = FALSE}
tccq_compile(f, backend = tccq_backend_source())  # return emitted C, do not compile
tccq_compile(f, backend = tccq_backend_tinycc())  # compile in memory via Rtinycc (default)
tccq_compile(f, backend = tccq_backend_shlib())   # compile via R CMD SHLIB, load via .Call()
```

The shared-library route is in the same space as
[`callme`](https://github.com/coolbutuseless/callme). Before compiling,
`tccq_compile()` validates backend capabilities against the target, the compile
context, and any explicit boundary APIs the module uses.

## A larger example: Viterbi

This follows the shape of the classic `quickr` Viterbi example while keeping the
declared-subset contract. It exercises braced declarations, matrix row/column
views, matrix writes, local vector fills, nested `for` loops, `max()`, and
`which.max()`.

```{r}
viterbi <- function(observations, states, initial_probs,
                    transition_probs, emission_probs) {
  declare({
    type(observations = integer(num_steps))
    type(states = integer(num_states))
    type(initial_probs = double(num_states))
    type(transition_probs = double(num_states, num_states))
    type(emission_probs = double(num_states, num_obs))
  })

  num_states <- length(states)
  num_steps <- length(observations)

  trellis <- matrix(0, nrow = length(states), ncol = length(observations))
  backpointer <- matrix(0L, nrow = length(states), ncol = length(observations))

  trellis[, 1] <- initial_probs * emission_probs[, observations[1]]

  for (step in 2:num_steps) {
    for (current_state in 1:num_states) {
      probabilities <- trellis[, step - 1] * transition_probs[, current_state]
      trellis[current_state, step] <- max(probabilities) *
        emission_probs[current_state, observations[step]]
      backpointer[current_state, step] <- which.max(probabilities)
    }
  }

  path <- integer(length(observations))
  path[num_steps] <- which.max(trellis[, num_steps])
  for (step in seq((num_steps - 1), 1)) {
    path[step] <- backpointer[path[step + 1], step + 1]
  }

  states[path]
}
```

Compile once, then call the compiled closure like a normal R function and check
it against the interpreter:

```{r}
set.seed(42)
num_steps <- 50L; num_states <- 6L; num_obs <- 20L
observations <- sample.int(num_obs, num_steps, replace = TRUE)
states <- seq_len(num_states)
initial_probs <- runif(num_states); initial_probs <- initial_probs / sum(initial_probs)
transition_probs <- matrix(runif(num_states^2), nrow = num_states)
transition_probs <- transition_probs / rowSums(transition_probs)
emission_probs <- matrix(runif(num_states * num_obs), nrow = num_states)
emission_probs <- emission_probs / rowSums(emission_probs)

compiled_viterbi <- tccq_compile(viterbi)
args <- list(observations, states, initial_probs, transition_probs, emission_probs)
identical(do.call(compiled_viterbi, args), do.call(viterbi, args))
```

Excluding compile time, the compiled closure is markedly faster than the
interpreter:

```{r, eval = requireNamespace("bench", quietly = TRUE)}
bench::mark(
  R = do.call(viterbi, args),
  tccquickr = do.call(compiled_viterbi, args),
  iterations = 100, check = FALSE
)[, c("expression", "median", "itr/sec", "mem_alloc")]
```

## Where to go next

- `vignette("the-r-subset")` — the precise accepted subset, the boundary model,
  and the optimization passes.
- `tccq_compile(f, mode = "ir")` — inspect the typed IR for any kernel.
- The architecture decisions live in `docs/decisions/` on GitHub.
