---
title: "The R subset, its boundaries, and the optimizations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The R subset, its boundaries, and the optimizations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tccquickr)
```

`tccquickr` compiles a deliberately narrow, **declared** subset of R. This
vignette states that subset precisely: the type language you declare, which
syntactic constructs are accepted (and what "accepted" means), where the
boundaries to the R interpreter are, and which optimizations the middle-end
performs.

## The declared type language

Every compiled function begins with `declare(type(...))`. A declared type is an
*element mode* at a *rank*:

| Rank | How you write it | Meaning |
|------|------------------|---------|
| scalar (0) | `double()`, `integer()`, `logical()` | a length-1 value |
| vector (1) | `double(NA)`, `integer(NA)`, `double(n)` | a vector; `NA` = unknown length, a name = a symbolic extent |
| matrix (2) | `double(nrow, ncol)`, `double(r, c)` | a 2-D array; bounds may be symbolic |

Element modes are `double`, `integer`, and `logical` (plus the internal `xlen`
index type). Symbolic extents (`n`, `num_states`, ...) tie argument shapes
together and feed shape/domain reasoning; they are how the Viterbi example in
`vignette("getting-started")` relates its matrices and vectors.

```{r}
# scalar in, vector in, matrix in — all declared
f <- function(x, v, m) {
  declare(type(x = double(NA), v = double(), m = double(3, 3)))
  x * v
}
```

## What "accepted" means: three dispositions

Each R construct has one of three dispositions, and the boundary between them is
the spec boundary of the compiler:

- **core** — compiles to a pure-C kernel with no R C API calls in the hot path.
- **boundary** — not expressible as a pure kernel, but lowered (under
  `fallback = "auto"`) to an explicit `r_eval` boundary node that calls back into
  R. Rejected under the default `fallback = "hard"`.
- **rejected** — not accepted in either mode (an honest gap or out of scope).

This table is **computed by probing the actual compiler**, so it cannot drift
from the implementation. It is grounded in R's own grammar, the yacc grammar
[`src/main/gram.y`](https://github.com/r-devel/r-svn/blob/master/src/main/gram.y).

```{r, echo = FALSE}
cov <- tccquickr:::tccq_grammar_coverage()
ord <- order(match(cov$disposition, c("core", "boundary", "rejected")), cov$production)
show <- cov[ord, c("production", "example", "disposition")]
names(show) <- c("Construct", "Probe", "Disposition")
knitr::kable(show, row.names = FALSE)
```

A drop in this coverage, or a `core` construct silently becoming a `boundary`,
is caught by `inst/tinytest/test_tccq_grammar.R`.

## The semantic model

The subset is small on purpose, and the rules are explicit so they live in the
IR and middle-end rather than appearing accidentally in the C emitter:

- `a <- expr` is a **local binding**, not mutation.
- `x[i]` is a checked, 1-based **indexed read**; `x[idx]` is a gather.
- `x[lo:hi]` is a contiguous **slice/view** (an explicit `view1` node), so a
  reduction over a slice need not copy first.
- `a[i] <- v` and `a[lo:hi] <- v` are **mutation barriers** on local storage.
- Direct mutation of a **formal argument** (e.g. `x[1] <- v` on an input) is
  rejected — inputs are borrowed.

```{r, error = TRUE}
direct_formal_mutation <- function(x, v) {
  declare(type(x = double(NA)), type(v = double()))
  x[1] <- v
  x
}
tccq_compile(direct_formal_mutation, mode = "code")
```

Writing through a local binding is fine: `y <- x` starts as an alias to the
input, and the first indexed write forces materialization into owned storage.

```{r}
assign_kernel <- function(x, i, v) {
  declare(type(x = double(NA)), type(i = integer()), type(v = double()))
  y <- x
  y[i] <- v
  y
}
unname(tccq_compile(assign_kernel)(c(1, 2, 3), 2L, 10))
```

## The boundary model

Optimization is organized around a **three-region split**, and the regions are
explicit in the IR rather than implied by codegen:

1. **Pure-C kernel region** — all non-boundary computation: plain C math and
   loops, no `Rf_*` calls. This is the only region targeted by optimization.
2. **C-API wrapper region** — argument/type checks and boundary-call plumbing,
   emitted with the R C API (`TYPEOF`, `XLENGTH`, `PROTECT`, ...).
3. **Boundary region** — every unsupported or dynamic call, represented as an
   explicit `boundary_call` node (currently backed by `r_eval`). It is a legality
   barrier: optimization never reorders or fuses across it.

Unsupported calls do not vanish into target special cases. Under the default
`fallback = "hard"` they are an error; under `fallback = "auto"` they become an
inspectable boundary node:

```{r}
fallback_kernel <- function(x) {
  declare(type(x = double()))
  floor(x)
}
ir <- tccq_compile(fallback_kernel, mode = "ir", fallback = "auto")
ir$ir$result$tag
ir$boundary_diagnostics[[1]]$original_call
```

## The optimization pipeline

The frontend lowers declared R into typed IR; the middle-end runs an ordered
pass pipeline; the target prints C; a backend compiles it. The default passes
(introspected live):

```{r, echo = FALSE}
passes <- vapply(tccquickr:::tccq_default_passes(), function(p) p$name, character(1))
desc <- c(
  validate_ir      = "Structurally validate the typed IR.",
  effects          = "Classify expression effects (pure vs boundary-bearing).",
  index_normalize  = "Normalize index/view access chains into a canonical form.",
  shape_domains    = "Derive shape/domain facts for vector expressions and locals.",
  kernelize        = "Rewrite expressions into explicit kernel IR (producer / materialize / fold / scalar_kernel).",
  fusion           = "Fuse legal producer/fold patterns (e.g. fold over a materialized producer becomes a fold over the producer).",
  boundary_collect = "Gather explicit r_eval boundary nodes.",
  boundary_validate= "Validate boundary-node legality and the R APIs they use.",
  storage_plan     = "Decide ownership: borrowed formals, owned locals, aliases, views, and materialization points.",
  allocation_plan  = "Plan allocations for outputs and temporaries.",
  protect_plan     = "Plan PROTECT/UNPROTECT for the C-API wrapper region."
)
knitr::kable(
  data.frame(pass = passes, does = desc[passes], row.names = NULL),
  col.names = c("Pass", "What it does")
)
```

The kernel IR concepts are what make the high-value decisions explicit:

- **`producer`** — an element computation over an index domain.
- **`materialize(producer)`** — allocate an output and write producer elements.
- **`fold(op, domain, elem)`** — reduce an element expression over a domain.

Fusion is therefore an IR rewrite, not an emitter accident. For
`sum((sin(x) + y) * y)`, `kernelize` first represents the reduction as a fold
over a *materialized* producer; `fusion` rewrites it to a fold over the producer
directly:

```{r}
mod <- tccquickr:::tccq_lower_module(tccquickr:::tccq_frontend(function(x, y) {
  declare(type(x = double(NA), y = double(NA)))
  sum((sin(x) + y) * y)
}))
base_passes <- list(
  tccquickr:::tccq_pass_validate_ir(), tccquickr:::tccq_pass_effects(),
  tccquickr:::tccq_pass_index_normalize(), tccquickr:::tccq_pass_shape_domains(),
  tccquickr:::tccq_pass_kernelize()
)
after_kernelize <- tccquickr:::tccq_run_passes(mod, base_passes)
after_fusion <- tccquickr:::tccq_run_passes(mod, c(base_passes, list(tccquickr:::tccq_pass_fusion())))
after_kernelize
after_fusion
```

## The longer view

The current target is C, but the pipeline is built so the high-value decisions
(types, ranks, ownership, shape domains, the elementwise domain, boundaries) are
made *before* any target syntax. A lowered IR seam sits between the middle-end
and the printers so the same decisions can drive other targets later
(Fortran/Rust/devices). The architecture and verification strategy — including
the R-as-oracle conformance number and the Lean transformation-soundness proofs
— are recorded in the decision records under `docs/decisions/` on GitHub.
